% Analysis 

```{r global_options, include = FALSE}
knitr::opts_chunk$set(cache = TRUE,
                      comment = "##",
                      collapse = TRUE,
                      warning = FALSE,
                      message = FALSE)

```


# Libraries

To reproduce the examples of this material, the R packages the following packages are needed.

```{r warning=FALSE, message=FALSE}

library(tidyverse)
library(EnvRtype)
library(metan)
library(rio)
library(factoextra)
library(FactoMineR)
library(ggrepel)
library(sp)
library(superheat)
library(corrr)

my_theme <- 
  theme_bw() +
  theme(panel.spacing = unit(0, "cm"),
        panel.grid = element_blank(),
        legend.position = "bottom")
```


# Datasets
## Traits
```{r}
me1 <- c("E3", "E4", "E5")
me2 <- c("E1", "E2", "E7")
me3 <- c("E6")
me4 <- c("E8")

df_traits <- 
  import("https://bit.ly/df_traits_wheat") |> 
  as_factor(1:6) |> 
  mutate(me = case_when(ENV %in% me1 ~ "ME1",
                        ENV %in% me2 ~ "ME2",
                        ENV %in% me3 ~ "ME3",
                        ENV %in% me4 ~ "ME4"))

```


## Climate variables
### Scripts to gather data

```{r eval=FALSE}
df_env <- import("https://bit.ly/loc_info_wheat")
ENV <- df_env$ENV |> as_character()
LAT <- df_env$LAT
LON <- df_env$LON
ALT <- df_env$ALT
START <- df_env$START
END <- df_env$END

# see more at https://github.com/allogamous/EnvRtype
df_climate <-
  get_weather(env.id = ENV,
              lat = LAT,
              lon = LON,
              start.day = START,
              end.day = END)
# GDD: Growing Degree Day (oC/day)
# FRUE: Effect of temperature on radiation use efficiency (from 0 to 1)
# T2M_RANGE: Daily Temperature Range (oC day)
# SPV: Slope of saturation vapour pressure curve (kPa.Celsius)
# VPD: Vapour pressure deficit (kPa)
# ETP: Potential Evapotranspiration (mm.day)
# PEPT: Deficit by Precipitation (mm.day)
# n: Actual duration of sunshine (hour)
# N: Daylight hours (hour)
# RTA: Extraterrestrial radiation (MJ/m^2/day)
# SRAD: Solar radiation (MJ/m^2/day)
# T2M: Temperature at 2 Meters
# T2M_MAX: Maximum Temperature at 2 Meters
# T2M_MIN: Minimum Temperature at 2 Meters
# PRECTOT: Precipitation
# WS2M: Wind Speed at 2 Meters
# RH2M: Relative Humidity at 2 Meters
# T2MDEW: Dew/Frost Point at 2 Meters
# ALLSKY_SFC_LW_DWN: Downward Thermal Infrared (Longwave) Radiative Flux
# ALLSKY_SFC_SW_DWN: All Sky Insolation Incident on a Horizontal Surface
# ALLSKY_TOA_SW_DWN: Top-of-atmosphere Insolation
# [1] "env"       "ETP"       "GDD"         "PETP"      "RH2M"      "SPV"
# [8] "T2M"       "T2M_MAX"   "T2M_MIN"   "T2M_RANGE" "T2MDEW"    "VPD"

# Compute other parameters
env_data <-
  df_climate %>%
  as.data.frame() %>%
  param_temperature(Tbase1 = 5, # choose the base temperature here
                    Tbase2 = 28, # choose the base temperature here
                    merge = TRUE) %>%
  param_atmospheric(merge = TRUE) %>%
  param_radiation(merge = TRUE)
```

### Tidy climate data
```{r}
env_data <- import("https://bit.ly/df_climate_tidy")
str(env_data)
id_var <- names(env_data)[10:23]
```





# Scripts
## Deviance analysis
### Model
```{r}
mod <-
  waasb(df_traits,
        env = ENV,
        gen = LINHAGEM,
        rep = BLOCO,
        resp = GY,
        wresp = 65) # maior peso para performance

waasb_env <-
  mod$GY$model %>%
  select_cols(type, Code, Y, WAASB) %>%
  subset(type == "ENV") %>%
  remove_cols(type) %>%
  rename(env = Code)
```

### BLUPs
```{r}
blupge <- gmd(mod, "blupge") 

blupge |> 
  make_mat(GEN, ENV, GY) |> 
  kableExtra::kable()
```

### BLUP-based stability
```{r}
indexes <- blup_indexes(mod)
kableExtra::kable(indexes$GY)
```


### GGE
```{r fig.width=12, fig.height=9}
mod_gge <- gge(blupge, ENV, GEN, GY, svp = 1)
p1 <- 
  plot(mod_gge,
       size.text.gen = 2.5,
       size.text.env = 2.5) +
  my_theme

p2 <- 
  plot(mod_gge,
       type = 2,
       size.text.gen = 2.5,
       size.text.env = 2.5) +
  my_theme

p3 <- 
  plot(mod_gge,
       type = 3,
       size.text.gen = 2.5,
       size.text.env = 2.5,
       size.text.win = 3.5) +
  my_theme

p4 <-
  plot(mod_gge,
       type = 4,
       size.text.gen = 2.5,
       size.text.env = 2.5,
       size.text.win = 3.5) +
  my_theme

p5 <-
  plot(mod_gge,
       type = 6,
       size.text.gen = 2.5,
       size.text.env = 2.5,
       size.text.win = 3.5) +
  my_theme

p6 <-
  plot(mod_gge,
       type = 8,
       size.text.gen = 2.5,
       size.text.env = 2.5,
       size.text.win = 3.5) +
  my_theme

arrange_ggplot(p1, p2, p3, p4, p5, p6,
               ncol = 3,
               tag_levels = "a",
               guides = "collect")

ggsave("figs/fig5_gge.png",
       width = 12,
       height = 9)
```


### WAASBY
```{r}
plot_waasby(mod, size.tex.lab = 6) +
  my_theme +
  theme(legend.position = c(0.8, 0.1))
ggsave("figs/fig6_waasby.png",
       width = 5,
       height = 6)
```


## Correlation between climate variables
```{r}
env_data |> 
  select_cols(tmean:rta) |> 
  correlate() |> 
  network_plot() +
  guides(color = guide_colorbar(barheight = 1,
                                barwidth = 20,
                                ticks.colour = "black")) +
  theme(legend.position = "bottom")

ggsave("figs/fig_network.png", width = 8, height = 8)
```


## Environmental kinships
```{r}

EC <- W_matrix(env.data = env_data,
               by.interval = TRUE,
               statistic = 'quantile',
               time.window = c(0, 30, 55, 70, 95, 130))


distances <-
  env_kernel(env.data = EC,
             gaussian = TRUE)

d <-
  superheat(distances[[2]],
            heat.pal = c("#b35806", "white", "#542788"),
            pretty.order.rows = TRUE,
            pretty.order.cols = TRUE,
            col.dendrogram = TRUE,
            legend.width = 2,
            left.label.size = 0.1,
            bottom.label.text.size = 5,
            bottom.label.size = 0.2,
            bottom.label.text.angle = 90,
            legend.text.size = 17,
            heat.lim = c(0, 1),
            padding = 1,
            legend.height=0.2)


ggsave(filename = "figs/fig2_heat_env.png",
       plot = d$plot,
       width = 6,
       height = 6)

```



## Principal Component Analysis
```{r }

prec <-
  env_data %>%
  remove_cols(LON:YYYYMMDD, daysFromStart) |>
  group_by(env) %>%
  summarise(prec = sum(prec))

# compute the mean by environment and year
df_long <-
  env_data %>%
  remove_cols(LON:YYYYMMDD, daysFromStart) |>
  remove_cols(prec) %>%
  pivot_longer(-env)

# bind environment WAASB, GY, and climate traits

pca <-
  df_long %>%
  means_by(env, name) %>%
  pivot_wider(names_from = name, values_from = value) %>%
  left_join(waasb_env |> rename(GY = Y), by = "env") %>%
  left_join(prec, by = "env") %>%
  mutate(me = case_when(env %in% me1 ~ "ME1",
                        env %in% me2 ~ "ME2",
                        env %in% me3 ~ "ME3",
                        env %in% me4 ~ "ME4")) |>
  column_to_rownames("env")


# compute the PCA with
pca_model <- PCA(pca,
                 quali.sup = 17,
                 graph = FALSE)


fviz_pca_biplot(pca_model,
                repel = TRUE,
                habillage = 17,
                # font.main = c(8, "bold", "red"),
                geom.var = c("arrow", "text"),
                title = NULL) +
  my_theme +
  theme(legend.title = element_blank())

ggsave("figs/fig3_pca.png", width = 4, height = 4)

```


## Environmental tipology {.panelset}
```{r}
names.window <-
  c('1-Tillering',
    '2-Boosting',
    '3-heading/flowering',
    '4-kernel milk stage',
    '5-physiological maturity',
    "")

out <-
  env_typing(env.data = env_data,
             env.id = "env",
             var.id = c("tmax", "vpd", "rta", "etp", "rh"),
             by.interval = TRUE,
             time.window = c(0, 30, 55, 70, 95, 130),
             names.window = names.window)

out2 <-
  separate(out,
           env.variable,
           into = c("var", "freq"),
           sep = "_",
           extra = "drop") |>
  mutate(me = case_when(env %in% me1 ~ "ME1",
                        env %in% me2 ~ "ME2",
                        env %in% me3 ~ "ME3",
                        env %in% me4 ~ "ME4"))

# plot the distribution of envirotypes
variable <- "tmax"
p1 <-
  out2 |>
  subset(var == variable) |> # change the variable here
  ggplot() +
  geom_bar(aes(x=Freq, y=env,fill=freq),
           position = "fill",
           stat = "identity",
           width = 1,
           color = "white",
           size=.2)+
  facet_grid(me~interval, scales = "free", space = "free")+
  scale_y_discrete(expand = c(0,0))+
  scale_x_continuous(expand = c(0,0))+
  xlab('Relative Frequency of the maximum air temperature (ºC)')+
  ylab("Environment")+
  labs(fill='Envirotype')+
  theme(axis.title = element_text(size=12),
        legend.text = element_text(size=9),
        strip.text = element_text(size=12),
        legend.title = element_text(size=12),
        strip.background = element_rect(fill="gray95",size=1),
        legend.position = 'bottom')

# by mega environment
p2 <-
  out2 |>
  subset(var == variable) |> # change the variable here
  sum_by(me, freq, interval) |>
  ggplot() +
  geom_bar(aes(x=Freq, y=me,fill=freq),
           position = "fill",
           stat = "identity",
           width = 1,
           color = "white",
           size=.2)+
  facet_wrap(~interval, nrow = 1)+
  scale_y_discrete(expand = c(0,0))+
  scale_x_continuous(expand = c(0,0))+
  xlab('Relative Frequency of the maximum air temperature (ºC)')+
  ylab("Mega Environment")+
  labs(fill='Envirotype')+
  theme(axis.title = element_text(size=12),
        legend.text = element_text(size=9),
        strip.text = element_text(size=12),
        legend.title = element_text(size=12),
        strip.background = element_rect(fill="gray95",size=1),
        legend.position = 'bottom') +
  scale_fill_discrete(direction = 1)

arrange_ggplot(p1, p2,
               heights = c(0.6, 0.4),
               tag_levels = "a",
               guides = "collect")
ggsave("figs/fig4_typology_tmax.png", width = 12, height = 7)
```



# Section info
```{r}
sessionInfo()
```

